Imperative Programming with Variational Effects

### 2018-08-11 paper/paper.tex
edit motivation

\section{Implementation of Formula Trees}

@@

In order to take advantage of the benefits of sharing provided by choices, we must develop variational
models of execution for a given language. We begin by showing that defining variational execution in a pure,
side-effect free evaluation context is a fairly straightforward exercise before turning to the much more
difficult problem of variational execution with side effects.

@@

If we do not have to worry about side effects in our language that we wish to introduce variation to then
our task is comparatively easy. We begin by considering a simple arithmetic expression language
defined by the following Haskell datatype:

@@

The process of ``variationalizing" this small example is quite straightforward. First, we add a
constructor to support choices in our language:

@@

Next we modify our interpreter. Instead of returning plain values of type \prog{Int},
we now return \emph{variational} values of type \prog{V Int}. We then take the cases
from the non-variational interpreter and make them variational by situating them within the
applicative functor instance for \prog{V} we defined in Section 2. Finally, we add a case
that handles our new \prog{AChc} constructor and we have completed the conversion of our
interpreter:

@@

To demonstrate why imperative programming with side effects requires us to rethink how we
variationalize a language and its execution, we turn to a small example in a simple imperative
language that supports basic control structures and variables with mutable state:

\begin{algorithmic}
\STATE $x \coloneqq 0$
\STATE $y \coloneqq$ getSecret()
\IF{$y$ is true}
\STATE{$x \coloneqq 1$}
\ENDIF
\end{algorithmic}

In a non-variational setting, we can reasonably expect the final state of the variables in this example
to be either $\{x \Rightarrow 0,\ y \Rightarrow \text{false}\}$ or
$\{x \Rightarrow 1,\ y \Rightarrow \text{true}\}$ depending on the value that
\prog{getSecret()} evaluates to. Now we consider the execution of the above example in a variational
setting. Suppose that the call to \prog{getSecret}
produces the variational value $\tagtree{A}{\text{true}}{\text{false}}$. We can view this as simply
combining the two possible non-variational execution paths described above into a
single execution varying in dimension $A$. As such, we would expect the final variable state
in this example to be  $\{x \Rightarrow \tagtree{A}{1}{0},\ y \Rightarrow \tagtree{A}{\text{true}}{\text{false}}\}$ by combining the final states above into choices over dimension $A$. We can verify this to be the proper variational result using the correctness relationship
specified in Section 2.

Of course, to arrive at the above result, we simply worked backwards from the non-variational
results. In order to take advantage of the benefits provided by sharing, we must instead develop a model
of variational execution that supports mutable state.  As we will see, the applicative functor instance for
\prog{V} is no longer sufficient by itself, necessitating a different approach to how we variationalize programs
that use mutable state.

@@

The problem comes in line 4, where we assign the variable $x$ in a conditional block. In the previous line, we check if the value of
$y$ is true. When $y$ is a variational value, allowing it to be both true or false in different variants, we only assign $x$ in those variants
in which $y$ is true. But if the state in our semantic domain is only a map, we have no way of passing the necessary information to the
assignment that it must only occur in particular variants and not in others. Our only choice is to assign the value of $x$ for all variants,
which would clearly be incorrect, as the assignment should only be carried out in the variant where $y$ is true. We must therefore
conclude that we need to carry this variational context information around in our state during execution in order to handle cases where
assignment occurs in a branch that is conditionally executed. Similar logic can be used to see that this context is also necessary to handle
variational lookup of variable values.

@@

%\begin{program}
%type Var = String
%data AExpr = N Int | Ref Var | GetSecret
%data BExpr = Equ AExpr AExpr
%data Stmt = Assn Var AExpr | If BExpr Block
%type Block = [Stmt]
%
%evalAExpr :: AExpr -> State -> Int
%evalAExpr (N i) _ = i
%evalAExpr (Ref x) s = lookup x s
%evalAExpr GetSecret = undefined
%
%evalBExpr :: BExpr -> State -> Bool
%evalBExpr (Equ l r) s = evalAExpr l s == evalAExpr r s
%
%evalStmt :: Stmt -> State -> State
%evalStmt (Assn x e) s = update x (evalAExpr e) s
%evalStmt (If e ss) s | evalBExpr e s = evalBlock ss s
%                     | otherwise     = s
% 
%evalBlock :: Block -> State -> State
%evalBlock b s = foldl (flip evalStmt) s b
%\end{program}

@@

\begin{algorithmic}
\STATE $y \coloneqq$ getSecret()
\IF{$y$ is true}
\STATE{\textbf{throw} $e$}
\ENDIF
\STATE{$x \coloneqq$ expensiveFn()}
\end{algorithmic}

In a non-variational context, the behavior of this program should be clear.
If the variable $y$ evaluates to \textbf{true}, then an error is thrown with value
$e$. At this point evaluation should stop, meaning that the variable $x$ is never
assigned in the final statement, avoiding the costly computation of \prog{expensiveFn}. This effect of halting execution and
returning an error value is the essence of exceptions in non-variational settings.

Now we consider the behavior of the same program in a variational context.
Suppose that the variable $y$ evaluates to the variational value
$\tagtree{A}{true}{false}$. This means that in the left alternative of $A$ we
would evaluate the body of the if statement, while ignoring it otherwise. Therefore,
in the left alternative of $A$ we throw an exception, but otherwise we continue on to the
final statement.

Clearly maintaining the same behavior from the non-variational setting in the variational setting
violates correctness. If we halt execution whenever an exception is
thrown in any variant, then we also stop the evaluation of variants that never encountered an error,
as in our example for the case $\neg A$. Correctness dictates that variants that never encountered
an exception should complete their execution uninhibited.

We also can't simply continue evaluation
in every variant regardless of whether or not we have encountered an error.
Suppose that calling \texttt{expensiveFn} normally would evaluate to the variational value $\tagtree{A}{1}{2}$, but at considerable computational cost in both alternatives.
Because we know that the left alternative of $A$ will ultimately evaluate to the thrown exception $e$, we would
like to avoid the cost involved in computing the value $1$ that we will just throw away later, mirroring how throwing an exception short-circuits evaluation in the
non-variational setting.

@@

\section{Comparison with Existing Work}

At this point, it should be apparent that introducing choices into a side-effectful imperative language
is anything but straightforward. The issues described above have been encountered repeatedly in efforts to
make variational versions of the While language \cite{varwhile}, PHP \cite{varex}, and Java \cite{varexj}.
The work on Varex the variational PHP interpreter, in particular demonstrates useful strategies for dealing
with mutable state and exceptions in a variational language.
In our work on the Resource DSL we independently discovered many of the same techniques used by Varex
to deal with these issues.
This work builds on the methods of handling variational effects discovered in the efforts to variationalize the
above languages by providing a formal operational semantics for the IMP language, a simple, albeit Turing-complete, imperative
language. It is our hope that a formal operational semantics for a simple imperative language will allow for easy application
and extension to variational imperative languages of differing levels of complexity.

>>>>

\section{Implementation of Formula Trees}
\label{sec:genvimpl}

@@

In order to take advantage of the benefits of sharing provided by choices, we must develop a variational
model of execution for a given language. We begin by showing that defining variational execution in a pure,
side-effect free evaluation context is a fairly straightforward exercise before turning to the much more
difficult problem of variational execution with side effects.

@@

If our target language has no side effects then
our task is comparatively easy. We begin by considering a simple arithmetic expression language
defined by the following Haskell datatype:

@@

The process of ``variationalizing'' this small example is quite straightforward. First, we extend the syntax by adding a
constructor to support choices in our language:

@@

Next we modify our interpreter. Instead of returning plain values of type \prog{Int},
we now return \emph{variational} values of type \prog{V Int}. We then take the cases
from the non-variational interpreter and make them variational by situating them within the
applicative functor instance for \prog{V} we defined in Section \ref{sec:genvimpl}. Finally, we add a case
that handles our new \prog{AChc} constructor and we have completed the conversion of our
interpreter:

@@

To demonstrate why imperative programming with side effects requires us to rethink how we
variationalize a language and its execution, we turn to a small example in a simple imperative
language that supports basic control structures and mutable variable references:

\begin{samepage}
\begin{algorithmic}
\STATE $x \coloneqq 0$
\STATE $y \coloneqq$ getSecret()
\IF{$y$ is true}
\STATE{$x \coloneqq 1$}
\ENDIF
\end{algorithmic}
\end{samepage}

In a non-variational setting, we can reasonably expect the final state of the variables in this example
to be either $\{x \Rightarrow 0,\ y \Rightarrow \CCkeyw{false}\}$ or
$\{x \Rightarrow 1,\ y \Rightarrow \CCkeyw{true}\}$ depending on the value that
\prog{getSecret()} returns. Now we consider the execution of the above example in a variational
setting. Suppose that the call to \prog{getSecret}
produces the variational value $\tagtree{A}{\CCkeyw{true}}{\CCkeyw{false}}$. We can view this as simply
combining the two possible non-variational execution paths described above into a
single execution varying in dimension $A$. As such, we would expect the final variable state
in this example to be  $\{x \Rightarrow \tagtree{A}{1}{0},\ y \Rightarrow \tagtree{A}{\CCkeyw{true}}{\CCkeyw{false}}\}$ by combining the final states above into choices over dimension $A$. We can verify this to be the proper variational result using the variation-preserving property
specified in Section \ref{sec:props}.

Of course, to arrive at the above result, we simply worked backwards from the non-variational
results. This naive approach cannot scale to domains with a high number of dimensions because of the exponential
explosion of variants. In order to take advantage of the benefits provided by sharing, we must instead develop a model
of variational execution that supports mutable state.  As we will see, the applicative functor instance for
formula trees is no longer sufficient by itself, necessitating a different approach to how we variationalize programs
that use mutable state.

@@

The problem comes in line 4, where we assign the variable $x$ in a conditional block. In the previous line, we check if the value of
$y$ is $\CCkeyw{true}$. When $y$ is a variational value, allowing it to be both $\CCkeyw{true}$ or $\CCkeyw{false}$ in different variants, we only assign $x$ in those variants
in which $y$ is $\CCkeyw{true}$. But if the state in our semantic domain is only a map, we have no way of passing the necessary information to the
assignment that it must only occur in particular variants and not in others. Our only choice is to assign the value of $x$ for all variants,
which would clearly be incorrect, as the assignment should only be carried out in the variant where $y$ is $\CCkeyw{true}$. We must therefore
conclude that we need to carry this variational context information around in our state during execution in order to handle cases where
assignment occurs in a branch that is conditionally executed. Similar logic can be used to see that this context is also necessary to handle
variational lookup of variable values.

@@

@@

\begin{samepage}
\begin{algorithmic}
\STATE $y \coloneqq$ getSecret()
\IF{$y$ is true}
\STATE{\textbf{throw} $e$}
\ENDIF
\STATE{$x \coloneqq$ expensiveFn()}
\end{algorithmic}
\end{samepage}

In a non-variational context, the behavior of this program should be clear.
If the variable $y$ evaluates to $\CCkeyw{true}$, then an error is thrown with value
$e$. At this point evaluation should stop, meaning that the variable $x$ is never
assigned in the final statement, avoiding the costly computation of \prog{expensiveFn}. This effect of halting execution and
returning an error value is the essence of exceptions in non-variational settings.

Now we consider the behavior of the same program in a variational context.
Suppose that the variable $y$ evaluates to the variational value
$\tagtree{A}{\CCkeyw{true}}{\CCkeyw{false}}$. This means that in the left alternative of $A$ we
would evaluate the body of the $\CCkeyw{if}$ statement, while ignoring it otherwise. Therefore,
in the left alternative of $A$ we throw an exception, but otherwise we continue on to the
final statement.

Clearly maintaining the same behavior from the non-variational setting in the variational setting
violates variation-preservation. If we halt execution whenever an exception is
thrown in any variant, then we also stop the evaluation of variants that never encountered an error,
as in our example for the case $\neg A$. Variation-preservation dictates that variants that never encountered
an exception should complete their execution uninhibited.

We also don't want to simply continue evaluation
in every variant regardless of whether or not we have encountered an error.
Suppose that calling \texttt{expensiveFn} normally would evaluate to the variational value $\tagtree{A}{1}{2}$, but at considerable computational cost in both alternatives.
Because we know that the left alternative of $A$ will ultimately evaluate to the thrown exception $e$, we would
like to avoid the cost involved in computing the value $1$ that we will just throw away later, mirroring how throwing an exception short-circuits evaluation in the
non-variational setting.

@@

\section{Towards a formal semantics}

At this point, it should be apparent that introducing choices into a side-effectful imperative language
is anything but straightforward. These issues are not new, and have been encountered in previous
work on variationalizing imperative languages \cite{varwhile,varex,varexj}. Along the way a number of
techniques have been developed to address these issues. Our work synthesizes these techniques into
a single big-step operational semantics for a simple imperative language. Along the way we also develop
novel techniques of our own, particularly when it comes to catchable exceptions, which to our knowledge
have not been adequately explored in a variational setting.

It is our hope that a formal semantics for common side effects encountered in variational imperative programming
will serve as the base for future work in this area. By collecting and formalizing the techniques shared among variational
imperative programming languages, future researchers should be able to (1) more easily define and create variational
languages by extending our semantics to more complex languages and (2) more easily formalize desriable properties in these languages.
In Chapter \ref{ch:rdsl} we demonstrate one successful application
of these techniques in the form of the Resource DSL.
